---
title: "DS_proj2"
author: "Hana Akbarnejad"
date: "3/16/2020"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(viridis)
library(ggplot2)
library(readr)
library(caret)
library(splines)
library(mgcv)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

```{r data}

# excluding Columbia University from Data
college_data = read_csv("College.csv") %>% 
  janitor::clean_names() %>% 
  filter(college != "Columbia University") %>% 
  select(-college) %>% 
  select(outstate, everything())

# defining model matrix of response variables and y(outstate tuition)
x = model.matrix(outstate~.,college_data)[,-1]
y = college_data$outstate
```

### Part a

In this part, I will create scatter plots to show the relationship between reponse variable and covariates. I used *featureplot* from *caret package* to do this.
```{r part_a}

#scatterplot
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)
featurePlot(x, y, plot = "scatter", labels = c("","Y"),
type = c("p"), layout = c(4, 4))
```

describe the results......................

### Part b

In this part I will fit a **smoothing spline model** using *terminal* as the only predictor and *outstate* as the outcome.

For the degrees of freedom, I will use two methods and plot the resulting fits:

* Using a range of degrees of freedom

* Using generalized cross-validation (GCV) to obtain degrees of freedom

```{r}

# df using GCV
ss_fit2 = smooth.spline(college_data$terminal, college_data$outstate)
ss_fit2

# plotting the ss fit
terminal_lims = range(college_data$terminal)
terminal_grid = seq(from = terminal_lims[1],to = terminal_lims[2])

ss_pred = predict(ss_fit2, x = terminal_grid)
ss_pred_df = data.frame(pred = ss_pred$y, terminal = terminal_grid)


ggplot(data = college_data, aes(x = terminal, y = outstate)) +
geom_point(color = rgb(.2, .4, .2, .5)) +
geom_line(aes(x = terminal, y = pred), data = ss_pred_df,
color = rgb(.8, .1, .1, 1)) + theme_bw()
```

From the above model, we can observe that the degrees of freedom from GCV is `r round(ss_fit2$df, 3)`, with the smoothing parameter of `r round(ss_fit2$spar, 3)`, and $\lambda$ of `r round(ss_fit2$lambda, 3)`.


Describe the results obtained...................

```{r}

```

